---
title: "InSciOut R / SQLite Tutorial"
author: "Geoffrey Mégardon"
date: "3 March 2017"
highlight: pygment
output:
  html_document:
    toc: true
    toc_float: true
# output:
#   md_document:
#     variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Useful links to use Rmarkdown:
- [Markdown Basics](http://rmarkdown.rstudio.com/lesson-8.html)
- [Display tables](http://rmarkdown.rstudio.com/lesson-7.html)

Note that I used RMarkdown for making the tutorial, but you can just use the R code in a Rscript file when making your own analyses.
Rmardown is useful to make pretty HTML, Word document, PDF reports.

## Importing the data and useful libraries
Before to start any analyses, we need to import the data and the libraries that we want to use. Note that you need to install the libraries from ```Tools -> Install Packages``` before to be able to import them in code.

In R, it is really rare to use the 'base' functions (i.e. out-of-the-box functions) for data manipulation. The libraries ```dplyr``` and ```data.table``` got the reputations to be much more powerful and easy to use than the 'base' functions. 

The libraries `DBI` and `RSQLite` (inside `DBI`) are here to load the tables we got in the SQL database. You first need to connect to the database and them, you ask the database what table you want. It is similar as to open a CSV file.

```{r results='hide', message=FALSE, warning=FALSE} 
library(dplyr)
library(data.table)
library(knitr) # just to display pretty tables in Rmarkdown
library(DBI) # to open connection with SQL database
db <- dbConnect(RSQLite::SQLite(), dbname = "../database/InSciOut.db")

```

Note that Rstudio is autocompleting the path if you use ```TAB```.

Let us display the first line of the meta table.
```{r } 
kable(dbGetQuery(db, "SELECT * FROM Meta_table LIMIT 1"))
```


Note that `kable` is here to make the table pretty. Usually (outside Rmarkdown), you can just write:

```{r eval=FALSE}
dbGetQuery(db, "SELECT * FROM Meta_table LIMIT 10")
```


> **TIP: ** Note that you can have a quick look to any tables loaded in memory from the GUI. For instance, let put the metatable in a variable:

```{r }
meta.table <- dbGetQuery(db, "SELECT * FROM Meta_table")
```

Now, go in the tab `Environment` and double-click on the table/variable's name you want to have a look at (here that is `meta.table`).

After you used the database, you need to close the connection:
```{r}

dbDisconnect(db)
unlink("../database/InSciOut.db")

```
## Data Base Organisation

Find more details in the other tutorial, on SQL / SQLite.

## Analysis 125

We start with Analysis 125 because it is simpler than the analysis 100.

### What we want to do

We use the analysis explained in row 125 in ```Data Analysis.xls```.
We want to obtain a table that contains the columns:

| Sample^1^ | Institution^2^ | Reference^3^ | JA_Design^4^ | PR_Advise^5^ | News_Source^6^ | News_Advices^7^ | News_Exageration^8^ | PR_Exageration^9^ | 
|------:|-----|-----|-----|-----|-----|-----|-----|-----|

1. Sample:
: take column `Sample` when (Source == "News")
2. Institution:
: take column `Institution` when (Source == "News")
3. Reference (PR number):
: take column `Reference` when (Source == "News")
4. JA_Design:
:     take column `Design_Actual` when (Source == "Journal Article - Body")
5. PR_Advise:
:     take column `Advice_Code` when (Source == "Final Press Release")

> **!WARNING!** note that in the document you speak about the JA - Body but   you cite cell E40, which correspond to PR

6. News_Source
:      take column `Source` when (Source_Category == "News")
7. News_Advise
:      take column `Advice_Code` when (Source_Category == "News")
8. News_Exageration
: Operation on **this** table:
      + 1 if `News_Advice` > `PR_Advice`
      + 0 if `News_Advice` = `PR_Advice`
      + -1 if `News_Advice` < `PR_Advice`
9. PR_Exageration
:      to take from Analysis 100 (a100)

> **!WARNING!** note that this last column needs input from Analysis 100

### Import the Root Tables (Meta, PR, JA and News)

Before to start, it is important to note that the resulting table as one line per news paper!

Also, let us select the columns that we actually need from each Root Tables (e.g. we only need the column `Design_Actual` from JA).
```{r }

db <- dbConnect(RSQLite::SQLite(), dbname = "../database/InSciOut.db")
news_table <- dbGetQuery(db, "SELECT Reference, Source, Advice_Code FROM News_table")
JA_table <- dbGetQuery(db, "SELECT Reference, Design_Actual FROM JABody_table")
PR_table <- dbGetQuery(db, "SELECT Reference, Advice_Code, Advice Sample_Code FROM PR_table")
Meta_table <- dbGetQuery(db, "SELECT Reference, Sample FROM Meta_table")

# Display the results:
kable(head(news_table, 5))
kable(head(PR_table, 5))
kable(head(JA_table, 5))
```

> Note that we imported JA_table from JABody_table that is a `view` that I made in the SQL database. This `view` only returns the rows of the JA_table that come from the article's body.

The result table that we are looking to generate will have one line per news paper, so we can use `news_table` as a base for the result table, to which we will add new columns.

### Create a Table that Crosses Information between JA, PR and News tables

Let us include the PR's `Advice_code` in `news_table`.
First, we need to change the name of `Advice_Code` in `news_table` for avoiding to override it with the PR's `Advice_Code`. Let's also take the opportunity to rename `Source` to `News_Source` 
```{r}
# setnames is a function from dtplyr (or data.table)
setnames(news_table, 
         old = c('Advice_Code','Source'), 
         new = c('News_Advice_Code','News_Source'))
setnames(PR_table, 
         old = c('Advice_Code'), 
         new = c('PR_Advice_Code'))

# display
kable(head(news_table, 1))
kable(head(PR_table, 1))
```

Now, we would like to put the column `PR_Advice_Code` in `news_table`. Unfortunately, for one PR/JA there are several news, so that `PR_table` does not have the same number of rows as `news_table`. Fortunately, R, python and SQL allows us to merge tables that don't have the same number of rows.

We will **merge** `PR_table` and `news_table` based **on** their `Reference` column (also called the **key**, or merged **by**). We will also ask R to use the column `Reference` in `news_table` as a reference, so that it keeps the number of rows present in `news_table`, and repeat the content of `PR_table` to fill the gaps.
```{r}
# using the data.table library:
d125 <- merge(x = news_table, y = PR_table, by='Reference', all.x = TRUE)

# alternative way, using dplyr:
# d125 <- left_join(news_table, PR_table, by='Reference')

# display the first 10 rows
kable(head(d125, 10) )
```

See here that we create a new table `d125`. As `news_table` is given as `x` to the merge function, `all.x = TRUE` tells R to keeps only the `Reference`'s values present in `news_table`. 

> **Note: ** As `x = news_table` is on the left of the function call, SQL and python would called what we did a **left join** operation.  

Let us include the columns from `JA_table` now. Note that we'll do the merge operation on the `d125` so that we keep the results of the previous merge operation.

```{r}
setnames(JA_table, 
         old = c('Design_Actual'), 
         new = c('JA_Design'))
# note that we do the merge operation on the d125 now:
d125 <- merge(x = d125, y = JA_table, by='Reference', all.x = TRUE)


# display
kable(head(d125, 10))
```

### Add New Columns To Our Table based on its Current Columns

Finally, we would like to compute the variable `News_Exageration` that compared the columns `News_Advice_Code` and `PR_Advice_Code` of our table `d125`.

> Note the usage of `ifelse()` is simialr to `numpy.where()` in python.
Example:
```{r} 
ifelse(c(1, -3, -3, 1, -3) > 0, 'yes', 'no')
```


```{r}
isNewsGreater <- d125$News_Advice_Code > d125$PR_Advice_Code
d125$News_Exageration = ifelse(isNewsGreater, 1, -1)

isNewsEqual = d125$News_Advice_Code == d125$PR_Advice_Code
d125[isNewsEqual, 'News_Exageration'] <- 0
# you could also write it:
# results_table$News_Exageration[isNewsEqual] <- 0

# display the first 10 rows:
kable(head(d125, 10))
```

### Add New Columns based on Older Analysis

We just need to import the csv file of Analysis 100 (I made the analysis before hand), and merge it with the table d125.

```{r}

d100 <- read.csv("../analyses/d100.csv")
sub.d100 <- d100 %>%
            select(Reference, PR_Exageration)

d125 <- merge(x = d125, y = sub.d100, by='Reference', all.x = TRUE)

# display only the first news of each Reference, so it is easier to check
kable(d125 %>% group_by(Reference) %>% filter(row_number() == 1)) 

```

So we are done with Analysis 125! :) it was an easy one, the next one will be a bit more complicated.


## Analysis 100

### What we want to do

We use the analysis explained in row 100 in ```Data Analysis.xls```.
We want to obtain a table that contains the columns:

| Sample^1^ | Institution^2^ | Reference^3^ | JA_Design^4^ | JA_Actual_Sample^5^ | JA_Sample^6^ | PR_Sample^7^ | PR_Exageration^8^ | News_Uptake^9^ | Total_News^10^
|------:|-----|-----|-----|-----|-----|-----|-----|-----|-----|

We want one row per PR.

1. Sample:
: take column `Sample` when (Source == "PR")
2. Institution:
: take column `Institution` when (Source == "PR")
3. Reference (PR number):
: take column `Reference` when (Source == "PR")
4. JA_Design:
:     take column `Design_Actual` when (Source == "Journal Article - Body")
5. JA_Actual_Sample:
:     take column `Sample_Actual` when (Source == "Journal Article - Body")

> **!WARNING!** note that in the document you speak about the JA - Body but   you cite cell E40, which correspond to PR

6. JA_Sample
:      take minimum of `Sample_Code` between Source == "Journal Article - Body" and Source == "Journal Article - Title and Abstract". Or, when grouping by JA and Reference, we keep the minimum.

7. PR_Sample
:      take column `Sample_Code` when (Source_Category == "PR")

> **!WARNING!** you refer to G29 in the spreadsheet, but I think that you meant E28

8. PR_Exageration
: Operation on **this** table:
      + **IF PR_Sample is 4,** 
      the exaggeration/understated column should read **-99**
      + **IF PR_Sample is 1 OR 2 AND JA_Sample is 1;** 
      the exaggeration/understated column should read **0** (i.e. correct)
      + **IF PR_Sample is 1 OR 2 AND JA_Sample is 3;** 
      the exaggeration/understated column should read **1** (i.e. exaggerated)
      + **IF PR_Sample is 3 AND JA_Sample is 3;** 
      the exaggeration/understated column should read **0** (i.e. correct)
      + **IF PR_Sample is 3 AND JA_Sample is 1;** 
      the exaggeration/understated column should read **-1** (i.e. understated)

> **!WARNING!** wasn't there an easier way to encode your data? that seems very complicated. Why should `PR_Sample > JA_Sample` not work here?

9. News_Uptake
:      if there any news articles attached to the PR reference `Reference`

10. Total_News
:     if yes, how many news articles for each PR

### Get the Root Tables
Note that the process in Analysis 125 is explained in much more details.
Here that is the table `PR_table`that will be used as reference because the result (target) table should have one row per PR.
Note that we don't need anything from the `news_table` this time, except their reference number.

```{r }

db <- dbConnect(RSQLite::SQLite(), dbname = "../database/InSciOut.db")
news_table <- dbGetQuery(db, "SELECT Reference FROM News_table")
JA_table <- dbGetQuery(db, "SELECT Reference, Source, Design_Actual, Sample_Actual, Sample_Code FROM JA_table")
PR_table <- dbGetQuery(db, "SELECT Reference, Sample_Code FROM PR_table")
Meta_table <- dbGetQuery(db, "SELECT Reference, Sample FROM Meta_table")

```

Just to compare, with the CSV files framework that I presented last time, the code would look like:
```{r eval=F, results='hide', message=FALSE, warning=FALSE}
df <- read.csv("../database/test_bigtable.csv") 
news_table <- df[df$Source_Category == 'News', c('Reference'), drop=FALSE]
JA_table <- df[df$Source_Category == 'JA', c('Reference', 'Source', 'Design_Actual', 'Sample_Actual', 'Sample_Code')]
PR_table <- df[df$Source_Category == 'PR', c('Sample', 'Reference', 'Sample_Code')]

```

### Get Minimum Between Two Rows in JA_table  
We will use the following reference as a example for the minimum function:
```{r}
kable(JA_table[JA_table$Reference == 'test-03-15-048',])

```

We group the `JA_table` by `Reference`, and take the minimum of the column `Sample_Code` for each reference number. Then, we write the result directly in the `Sample_Code`.

```{r}
### Note that %>% is the syntax that dplyr uses to pipe/chain operations: 
JA_table <- JA_table %>% 
          group_by(Reference) %>%
          mutate(Sample_Code = min(Sample_Code)) # we override the column Sample_Code with its minimum

kable(JA_table[JA_table$Reference == 'test-03-15-048',])

```

It works!
So now we need to remove the `JA - Title and Abstract` rows.

``` {r }
# with dplyr's syntax:
JA_table <- JA_table %>%
            filter(Source == 'Journal Article - Body')

# with R base's syntax
# JA_table = df[df$Source == 'Journal Article - Body',]

kable(JA_table[JA_table$Reference == 'test-03-15-048',])

```

### Merge JA, PR and Meta tables

Let us merge the tables using the functions from the `data.table` library:
``` {r }
### Create a Table that Crosses Information between JA, PR and News tables ----
# merge the PR Table to the News table
setnames(JA_table, 
         old = c('Sample_Code'), 
         new = c('JA_Sample_Code'))
setnames(PR_table, 
         old = c('Sample_Code'), 
         new = c('PR_Sample_Code'))

```

```{r}
d100 <- merge(x = PR_table, y = JA_table, by='Reference')
d100 <- merge(x = d100, y = Meta_table, by='Reference')
d100 <- select(d100, -Source) # remove the column 'Source'
# note that you could also write:
# d100 <- d100 %>% select(-Source)
kable(d100[1:5,])
```

### Create a New Columns Based on Rules

We need to make the column `PR_Exageration`according to rules that combined `PR_Sample_Code` and `JA_Sample_Code`.
As those rules are long to write, we use the library `data.table` to shorten the code (and it will also speed up the computation).
```{r}
setDT(d100) # we make it a Data.Table, allows to go faster and write less

d100 <- 
  d100[(PR_Sample_Code %in% c(1,2)) & (JA_Sample_Code == 1), PR_Exageration:= 0 ]
d100 <-
  d100[(PR_Sample_Code %in% c(1,2)) & (JA_Sample_Code == 3), PR_Exageration:= 1 ]
d100 <-
  d100[(PR_Sample_Code == 3) & (JA_Sample_Code == 3), PR_Exageration:= 0 ]
d100 <-
  d100[(PR_Sample_Code == 3) & (JA_Sample_Code == 1), PR_Exageration:= -1 ]
d100 <-
  d100[(PR_Sample_Code == 4), PR_Exageration:= -99 ]

# use
# setDF(d100) # to come back to a Data.Frame (if needed)

# without the setDT(d100), you would need to write:
# d100$PR_Exageration[(d100$PR_Sample_Code %in% c(1,2)) & (d100$JA_Sample_Code == 1) ] = 0

# display (we need to add with=FALSE because we used a data.table)
kable(head(d100 %>% select(PR_Sample_Code, JA_Sample_Code, PR_Exageration), 10))
```

### Create Columns that counts the News for each PR

We are adding the columns that report the number of news per PR and the news uptakes (yes if number of news > 0) :
```{r}

news_table <- news_table %>% 
              group_by(Reference) %>%
              mutate(News_Uptake = 'yes', Total_News= n()) %>%
              filter(row_number()==1)
```

In the previous chunk of code, we grouped the `news_table` by Reference, and then:
- we put 'yes' in the new columns `News_Uptake` for all of these references.
- we counted the number of rows in each group and put the results in the new column `Total_News`. The `n()` is a dplyr function that returns the number of row of the current table/group.
- once done, we keep only the first row of each group with code `filter(row_number()==1)`. 

The filter operation is necessary so that we keep only one row per reference for the merge operation in next section. Otherwise the merge operation will return several rows for all the references present in `news_table`.

### Merge News Table to the Result Table

Here, for some exotism, we are using the `left_join` function to make the merge operation. `Merge` and `Join` operations are the same stuff.

``` {r}

d100 <- left_join(d100, news_table, by='Reference')

d100$News_Uptake[is.na(d100$News_Uptake)] <- 'no' # note that NA can only be detected with is.na()
d100$Total_News[is.na(d100$Total_News)] <- 0
kable(head(d100, 10))
```

Note that, by doing `all.x = TRUE`, I only kept references that have a PR/JA (the one in `d100`), so if there are news that have no related JA/PR, they won't appear.

Don't forget to close the connection:

```{r}
dbDisconnect(db)
unlink("../database/InSciOut.db")
```


