---
title: "Debug_Rachel_code"
author: "Geoffrey Mégardon"
date: "6 April 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(knitr)
library(dplyr)
```

```{r}
library(DBI)
getwd()
db <- dbConnect(RSQLite::SQLite(), dbname = "../../database/InSciOut.db")
```

## Bug report
Rachel:

> I attempted the two causal language tables (rows 73 and 146). I was fine with the first one including the journal article and press release but when I try the one containing the news articles I end up with 299 rows rather than 33. 
> I've been through the code and the error occurs whenever I pull data from the temporary table into the main table.
I'm guessing it's a joining issue but I can't figure it out.
> Code attached (note JAPR_language_73 is a view that I have created to get the data for whether / not the press release was exaggerated)

> I want as many rows as news articles and there are 33 news articles.
it's somehow duplicating the news articles. Eg. for the first press release there are 8 news articles but it's giving me 8 x Times articles, 8 x independent articles, 8 x mirror articles etc.

Attached: PN_NA_Language.sql
```{sql connection=db, eval=F}

WITH
tmpT1 AS
(
SELECT Reference,
    CASE 
    WHEN News_table.TMS_IVDV_Same = 1  THEN MAX(News_table.Title_Code, News_table.MS_Code)
    WHEN News_table.TMS_IVDV_Same = 0  THEN News_table.MS_Code
    END
    News_CaCode
    
    FROM News_table
    )
    
SELECT
Reference,
Meta_table.Institution, 
Meta_table.Sample,
JABody_table.Design_Actual, 
JAPR_language_73.JA_CaCode,
JAPR_language_73.PR_Exag_Adams,
JAPR_language_73.PR_Exag_BMJ,
JAPR_language_73.PR_ActExag_Adams,
News_table.Source as News_Source,
-- statement of cause from news
tmpT1.News_CaCode,

-- is the news exaggerated relative to the JA (Adams et al coding)

CASE
    WHEN JABody_table.Design_Actual IN (0,8)                                         THEN -99
    WHEN JAPR_language_73.JA_CaCode IN (0,1,-9)   OR tmpT1.News_CaCode IN (0,1,-9)   THEN -99
    
    WHEN  JAPR_language_73.JA_CaCode IN (2,3,4)   AND tmpT1.News_CaCode IN (2,3,4)   THEN 0
    WHEN  JAPR_language_73.JA_CaCode IN (5,6)     AND tmpT1.News_CaCode IN (2,3,4)   THEN -1
    WHEN  JAPR_language_73.JA_CaCode IN (2,3,4)   AND tmpT1.News_CaCode IN (5,6)     THEN 1
    WHEN  JAPR_language_73.JA_CaCode IN (5,6)     AND tmpT1.News_CaCode IN (5,6)     THEN 0

    END
    News_Exag_Adams,


-- is the news exaggerated relative to the JA (BMJ coding)

CASE
    WHEN JABody_table.Design_Actual IN (0,8)                                         THEN -99
    WHEN JAPR_language_73.JA_CaCode IN (0,1,-9)   OR tmpT1.News_CaCode IN (0,1,-9)   THEN -99

    WHEN  JAPR_language_73.JA_CaCode = tmpT1.News_CaCode                             THEN 0
    WHEN  JAPR_language_73.JA_CaCode < tmpT1.News_CaCode                             THEN 1
    WHEN  JAPR_language_73.JA_CaCode > tmpT1.News_CaCode                             THEN -1
    END      
    News_Exag_BMJ, -- CHECK THAT THIS IS CORRECT WITH SOL
    


--- is the news exaggerated relative to the study design (Adams et al coding)

CASE
   WHEN JABody_table.Design_Actual IN (0,8)                                        THEN -99
   WHEN tmpT1.News_CaCode IN (0,1,-9)                                              THEN -99
   WHEN JABody_table.Design_Actual IN (1,2,3,5,6) AND tmpT1.News_CaCode IN (2,3,4) THEN 0
   WHEN JABody_table.Design_Actual IN (4,7,9)     AND tmpT1.News_CaCode IN (2,3,4) THEN -1
   WHEN JABody_table.Design_Actual IN (1,2,3,5,6) AND tmpT1.News_CaCode IN (5,6)   THEN 1
   WHEN JABody_table.Design_Actual IN (4,7,9)     AND tmpT1.News_CaCode IN (5,6)   THEN 0
   END
   News_ActExag_Adams
   
FROM
News_table
LEFT JOIN Meta_table USING(Reference)
LEFT JOIN JABody_table USING(Reference)
LEFT JOIN JAPR_language_73 USING(Reference)
LEFT JOIN tmpT1 USING(Reference)

```


## Fisrt thought

there might be a duplicate in JABody (note that I don' t have the same database here):
```{sql connection=db}
WITH
T1 AS (
SELECT Reference, JABody_table.Source
FROM JABody_table
)

SELECT Reference, COUNT(*)
FROM T1
GROUP BY Reference
HAVING COUNT(*)>1

```

Such a duplicate can increase the number of duplicates x2 for each JOIN operation:
```{sql connection=db}
WITH
T1 AS (
SELECT Reference, JABody_table.Source
FROM JABody_table
LEFT JOIN JATitle_table USING(Reference)
LEFT JOIN PR_table USING(Reference)
)

SELECT Reference, COUNT(*)
FROM T1
GROUP BY Reference
HAVING COUNT(*)>1

```

## Number of News in our Database

```{sql connection=db}
SELECT
COUNT()

FROM
News_table

```

We got 401 News.

## Mock Query

We don' t have access to the view JAPR_language_73 and the error seems to only involve the temporary table.

> I've been through the code and the error occurs whenever I pull data from the temporary table into the main table.

Adapted from PN_NA_Language.sql
```{sql connection=db}

WITH
tmpT1 AS
(
SELECT Reference,
    CASE 
    WHEN News_table.TMS_IVDV_Same = 1  THEN MAX(News_table.Title_Code, News_table.MS_Code)
    WHEN News_table.TMS_IVDV_Same = 0  THEN News_table.MS_Code
    END
    News_CaCode
    
    FROM News_table
    )
    
SELECT
Reference,
Meta_table.Institution, 
Meta_table.Sample,
JABody_table.Design_Actual, 
News_table.Source as News_Source,
-- statement of cause from news
tmpT1.News_CaCode,

-- is the news exaggerated relative to the JA (Adams et al coding)

CASE
    WHEN JABody_table.Design_Actual IN (0,8) THEN -99
    END
    News_Exag_Adams,


-- is the news exaggerated relative to the JA (BMJ coding)

CASE
    WHEN JABody_table.Design_Actual IN (0,8) THEN -99
    END      
    News_Exag_BMJ, -- CHECK THAT THIS IS CORRECT WITH SOL
    
CASE
   WHEN JABody_table.Design_Actual IN (0,8)                                        THEN -99
   WHEN tmpT1.News_CaCode IN (0,1,-9)                                              THEN -99
   WHEN JABody_table.Design_Actual IN (1,2,3,5,6) AND tmpT1.News_CaCode IN (2,3,4) THEN 0
   WHEN JABody_table.Design_Actual IN (4,7,9)     AND tmpT1.News_CaCode IN (2,3,4) THEN -1
   WHEN JABody_table.Design_Actual IN (1,2,3,5,6) AND tmpT1.News_CaCode IN (5,6)   THEN 1
   WHEN JABody_table.Design_Actual IN (4,7,9)     AND tmpT1.News_CaCode IN (5,6)   THEN 0
   END
   News_ActExag_Adams
   
FROM
News_table
LEFT JOIN Meta_table USING(Reference)
LEFT JOIN JABody_table USING(Reference)
LEFT JOIN tmpT1 USING(Reference)

```


We get a number of rows > 1000 while there are only 401 News.
Now that we see that the bug still occurs while removing the view JAPR_language_73, we can actually remove even all the rows and complicated things around and only return the number of rows:

```{sql connection=db}

WITH
tmpT1 AS
(
SELECT Reference
FROM News_table
)
    
SELECT
COUNT()

FROM
News_table
LEFT JOIN Meta_table USING(Reference)
LEFT JOIN JABody_table USING(Reference)
LEFT JOIN tmpT1 USING(Reference)

```

## First Problem, the duplicate References:

As stated by Rachel to remove the Temporary table helps, we obtain 458 rows. But that is not exactly equal to the nb of News.
You can see how the number increases with the number of JOIN operations.

```{sql connection=db}

SELECT
COUNT(*)

FROM
News_table
LEFT JOIN PR_table USING(Reference)
LEFT JOIN Meta_table USING(Reference)
--LEFT JOIN JABody_table USING(Reference)


```

This is because there is a duplicate Reference in PR_table (you can try with JABody and JATitle tables as well):
```{sql connection=db}

SELECT Reference, COUNT(*)
FROM PR_table
-- FROM JABody_table
-- FROM JATitle_table
GROUP BY Reference
HAVING COUNT(*)>1

```

## Second Problem, Reference is not a unique ID for News

Ok, so we have 1 duplicate Reference in PR_table.
But, in News_table, there are even more duplicates of Reference:
```{sql connection=db}

SELECT Reference, COUNT(*)
FROM News_table
GROUP BY Reference
HAVING COUNT(*)>1

```

Yes, indeed, Reference is a unique ID only for PR, JABody and JATitle tables!

Now, this is not a problem when JOINing News_table with PR_table, JABody_table and JATitle_table because Reference is still unique in those tables.

However when `JOIN`ing tmpT1 and News_table, we need to note that tmpT1 is derivated from News_table: it contains the same rows of Reference.
When `JOIN`ing them `USING(Reference)`, SQL has no idea on how to relate the rows with each other, because Reference are not unique in any of them!

We need to add another columns to make sure that the key to related the two tables is unique for each row. `USING(Reference, Source)` should do the job:

```{sql connection=db}

WITH
tmpT1 AS
(
SELECT Reference, Source
FROM News_table
)
    
SELECT
COUNT()

FROM
News_table
--LEFT JOIN Meta_table USING(Reference)
--LEFT JOIN JABody_table USING(Reference)
LEFT JOIN tmpT1 USING(Reference, Source);

```

